---
title: "Bruce Campell NCSU ST 534 Exam 2"
date: "`r format(Sys.time(), '%d %B, %Y')`"
fontsize: 12pt
header-includes:
   - \usepackage{bbm}
output: pdf_document
---

---
```{r setup, include=FALSE,echo=FALSE}
rm(list = ls())
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(dev = 'pdf')
knitr::opts_chunk$set(cache=TRUE)
knitr::opts_chunk$set(tidy=TRUE)
knitr::opts_chunk$set(prompt=FALSE)
knitr::opts_chunk$set(fig.height=5)
knitr::opts_chunk$set(fig.width=7)
knitr::opts_chunk$set(warning=FALSE)
knitr::opts_chunk$set(message=FALSE)
knitr::opts_knit$set(root.dir = ".")
library(latex2exp)   
library(pander)
library(ggplot2)
library(ggplot2)
library(GGally)
```

##1 sunspotz analyis

Consider the series sunspotz in the package astsa and answer the following questions:

### (a) Plot sunspotz. Comment on the notable features of the time series? 

```{r}
rm(list = ls())
library(astsa)
data(sunspotz, package="astsa")
df <- sunspotz
plot(ts(df))
```

From the doscumentaion sunspotz. is the *biannual smoothed (12-month moving average) number of sunspots from June 1749 to December 1978; n = 459*.  The format is ```Time Series: Start = c(1749, 1) End = c(1978, 1) Frequency = 2```  There appear to be two or more fundamental frequencies in the series.  We note the sppearance of short term and long term cyclical behavior. In regards to modelling with this data we would want to take into consideration that the period of observations is shore realtive to the eveolution of the dynamical system generating the data, and that the technology for making observations may have evolved over the period of data collecation. 


### (b) Plot the ACF of sunspotz. Do you think it is appropriate to model this using a
stationary ARMA process ? 

```{r}
acf2(df)

```

We note the presense of strong long range correlation.  At first glance we would be hesitant to model this with a stationary ARMA process.  We'd possibly perform a unit root test and consider modelling this data with an ARIMA process. For fun, we do some modelling in the temporal doamin. 


```{r ,results='hide',fig.keep='all'}
invisible(model <-sarima(df,3,1,2))
model$ttable
```

We note some outlier in the the standardised residuals, heavy tails in the QQ plot, and that the individual Ljung-Box pp-vaules are all close to being significant.  We also note a presense of a lag 11,12 acf component in the rsiduals.  This is consistent with our intial observation that the data had short term and long term cyclic trends. 

For more fun let's calculate the Ljung-Box-Pierce Q-statistic to check for systemic autocorrelation in the residuals. 

```{r}

n <- length(df)
H <- 20
r <-model$fit$residuals[1:H]
acf.residuals <- acf(r,H, main="ACF of residuals")
sum.denominator <- n-seq(H, 1, by = -1)   
r.s <-  acf.residuals$acf^2 / sum.denominator
Q <- n*(n+2) *sum(r.s)
Q
```

We see based on the Q-statistic that we have significant correlation structure remaining in the residuals. 


### (c) Plot the periodogram of the sunspotz series using the spectrum command. 

We used the definition of the periodogram and the fft to calculate $I(\omega_j) = | d(\omega_j) |^2$ taking into account that we had to normalize the fft result, take out the DC component, and only use half of the fft.  We also plotted with the spec.pgram 

```{r}
df.periodogram <- Mod(fft(df-mean(df)))^2/length(df)   
df.half <-ts( df.periodogram[1:length(df)/2])
plot(df.half, main="Periodogram from fft")

df.periodogram.spec <- spec.pgram(df,taper = 0,log = "no",main="Periodogram from spec")
```
What we have plotted is the raw periodogram.  The raw periodogram is not a cosnistent estimator for the spectral density.  We could optinally use a smoothing kernel in the spec.pgram function to obtain a consistent estimate.

### (d) Find the maximum of the periodogram values and the frequency at which the
maximum occurs. 

We're interested in the max and the other peaks to try and understant the frequency components in the signal.  Here we extract the max.  The spec.pgram method returns a spectrum object with the data in ```freq``` , ```spec``` lists.  


```{r}
max.spec.loc <- which.max(df.periodogram.spec$spec)
max.spec <- df.periodogram.spec$spec[max.spec.loc]
max.spec.freq <- df.periodogram.spec$freq[max.spec.loc]

pander(data.frame(max.spec.freq = max.spec.freq,max.spec=max.spec))
```
We tried to fix up some code to fnd locations of the other peaks in the periodogram.  The idea is to use overlapping intervals to fund the local peaks via a spline maximization. It doesn't work yet :(

We did try to smooth the spectrum with a Danniel Kernel first. We'll keep that plot for reference. Note the 2 periodic componets are highlighted better in the Danniel smoothed spectrum.

```{r}
k = kernel("daniell", 4)
df.periodogram.spec.danniel <- spec.pgram(df,k,taper = 0,log = "no",main="Periodogram from spec with Danniel kernel smoothing.")
# f <- function(x, q, d) spline(q, d, xout = x)$y
# x <- df.periodogram.spec.danniel$freq
# y <- df.periodogram.spec.danniel$spec
# nb <- 20 # choose number of intervals
# iv <- embed(seq(floor(min(x)), ceiling(max(x)), len = nb), 2)[,c(2,1)]
# # make overlapping intervals to avoid problems if the peak is close to 
# # the ends of the intervals (two modes could be found in each interval)
# iv[-1,1] <- iv[-nrow(iv),2] - 2
# # The function "f" is maximized at each of these intervals
# iv
# # the solution is a local maximum
# gr.thr <- 0.01
# hes.thr <- 0.03i
# 
# require("numDeriv")
# vals <- matrix(nrow = nrow(iv), ncol = 3)
# grd <- hes <- rep(NA, nrow(vals))
# for (j in seq(1, nrow(iv)))
# {
#   opt <- optimize(f = f, maximum = TRUE, interval = iv[j,], q = x, d = y)
#   vals[j,1] <- opt$max
#   vals[j,3] <- exp(opt$obj)
#   grd[j] <- grad(func = f, x = vals[j,1], q = x, d = y)
#   hes[j] <- hessian(func = f, x = vals[j,1], q = x, d = y)
#   if (abs(grd[j]) < gr.thr && abs(hes[j]) > hes.thr)
#     vals[j,2] <- 1
# }
# # bin the peaks to avoid similar local maxima
# vals[,1] <- round(vals[,1], 2)
# if (anyNA(vals[,2])) {
#   peaks <- unique(vals[-which(is.na(vals[,2])),1])
# } else 
#   peaks <- unique(vals[,1])
# 
# plot(df.periodogram.spec$freq, df.periodogram.spec$spec, log = "y", type = "l")
# abline(v = peaks, lty = 2)
```



### (e) Find an AR-based estimate of the spectral density of sunspotz using the com-
mand spec.ar. 

First we note the caution from the documentaiton of spec.ar 
*Warning Some authors, for example Thomson (1990), warn strongly that AR spectra can be misleading.* 
We'll keep this in mind as we proceed. 

```{r}
df.spectrum.ar <- spec.ar(df,log="no")#, log="no", ylim=c(0,.5), plot = TRUE)
```

This looks similar to the Danniel smmothed spectrum only smoother. The second peak seems less distinct in this spectra. 


### (f) Find the maximum of the estimated spectral density and the frequency at which
the maximum occurs. 


```{r}
max.spec.loc <- which.max(df.spectrum.ar$spec)
max.spec <- df.spectrum.ar$spec[max.spec.loc]
max.spec.freq <- df.spectrum.ar$freq[max.spec.loc]

pander(data.frame(max.spec.freq = max.spec.freq,max.spec=max.spec))
```

### (g) Find the two estimates of the period of the most prominent periodic component
of the sunspotz series using your answers from parts (d) and (f). What does it
say about the Sunspot cycle? 

I'm not sure what we're being asked in this part.  The period is given by $\frac{1}{\omega}$ so the max of the ar estimated spectral density has a period of $\frac{1}{0.09619}= $ ```r 1/0.09619```

The max of the spectrum given by the fft is $\frac{1}{0.09167}= $ ```r 1/0.09167```   So the period is likely between 10-11 years, but it's not clear right now how we would statistically estimate this. 

We can look at the frequency of the max of the Danniel smoothed spectrum for additional insight. 
```{r}
max.spec.loc <- which.max(df.periodogram.spec.danniel$spec)
max.spec <- df.periodogram.spec.danniel$spec[max.spec.loc]
max.spec.freq <- df.periodogram.spec.danniel$freq[max.spec.loc]
pander(data.frame(max.spec.freq = max.spec.freq,max.spec=max.spec))
```

We learned how to parametrically and non-parametrically estimate the spectrum $f(\omega_0)$ at a location $\omega_0$.  Maybe we'll return to this and perform those estimates. 


##2 
Suppose that $\{S_t\}$ and $\{N_t\}$ are zero mean SOS and independent time series with
ACvFs $\gamma_1()$ , $\gamma_2$ and and spectral densities $f_1(\omega)$ and $f_2(\omega)$, respectively. Let

$$X_t = S_t + A S_{t-D} + N_t$$

where $A \in (0,\infty)$ is a constant and $D > 0$ is an integer.

### (a) Find the ACvF of $\{X_t\}$.

### (b) Using part (a) or otherwise, show that the spectral density of fXtg is given by

$$f_x(\omega) = [1 + A^2 + 2A cos(2 \omegaD)]f_1(\omega) + f2(\omega)$$



## 3. Consider the unemployment time series unemp given in the package astsa.

### (a) Plot the unemp series and its ACF upto lag 50. 

```{r}
rm(list = ls())
library(astsa)
data(unemp, package="astsa")
df <- unemp
plot(ts(df))
acf2(df,50)
```


### (b) Fit a Seasonal ARIMA model of order $(0 1 2) \times (0,1,1)_{12}$ and present the R output.

```{r ,results='hide',fig.keep='all'}
invisible(model <-sarima(df,0,1,2,0,1,1,12))
model$ttable
```

For lag 1 the Ljung-Box statistic shows significant correlation in the residuals.  

Let's calculate the Ljung-Box-Pierce Q-statistic to check for systemic autocorrelation in the residuals.  We'll extract the residuals and do the calculation by hand - there's and R function for this ```Box-Test```  that we've experimented with.  we'll revisit this.

```{r}
n <- length(df)
H <- 20
r <-model$fit$residuals[1:H]
acf.residuals <- acf(r,H, main="ACF of residuals")
sum.denominator <- n-seq(H, 1, by = -1)   
r.s <-  acf.residuals$acf^2 / sum.denominator
Q <- n*(n+2) *sum(r.s)
Q
```


### (c) Consider the Ljung-Box statistic p-value plot and the ACF of the residuals. What
do these indicate about inadequacy of the fitted model ? 

Thhe Ljung-Box statistic shows nearly significant and autocorrelation in the residuals at some lags.  We would look into increasing the AR order for our model to reduce this. 

### (d) Refine the SARIMA model based on your answer to the last part and present the
R output for the new SARIMA model. Indiacte how the new model improves on
the model in part (b). 

```{r}
invisible(model <-sarima(df,2,1,2,1,1,1,12))
model$ttable
```

In looking at my calculation of the Ljung-Box-Pierce Q-statistic I'm suspicious of my code based on the plot above. REVISIT

```{r}
n <- length(df)
H <- 20
r <-model$fit$residuals[1:H]
acf.residuals <- acf(r,H, main="ACF of residuals")
sum.denominator <- n-seq(H, 1, by = -1)   
r.s <-  acf.residuals$acf^2 / sum.denominator
Q <- n*(n+2) *sum(r.s)
Q
```
### (e) Use the fitted model in part (d) to forecast the next 12-months.



